---
title: "DESeq2 and GOSeq"
author: "Ismael Gutierrez-Fernandez"
date: "2025-02-21"
output: html_document
---

## Original script by Philippa Borrill (8.21.2018, last updated 13.6.2022), with some minor modifications by Ismael Gutierrez-Fernandez

# AIM: RUN DESeq2 FOR COMPARING GENE EXPRESSION "control" vs "Treatment"

## STEP 0. Install and load the packages. Define directories and import the data

```{r}
# 0.1. Install. Run ONLY if the packages are not previously installed

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.18") # This installs version 3.18 of BiocManager that works on R version 4.3.2; for new versions change "version=" or check the source webpage: https://bioconductor.org/install/

BiocManager::install("DESeq2")
BiocManager::install("goseq")

```

```{r}
# 0.2. Load packages for this script.

library(DESeq2)
library(goseq)
library(dplyr)
library(tidyr)
library(ggplot2)

```

```{r}
# 0.3. Define paths

data_dir <- "//jic-hpc-data/hpc-home/rna-training/R/tximport/" # Define the path for the directory where the tximport results are placed

out_dir_DESeq2 <- "//jic-hpc-data/HPC-Home/rna-training/R/DESeq2/" # Define the path for the directory wehre DESeq2 results will be saved

out_dir_GO <- "//jic-hpc-data/HPC-Home/rna-training/R/DESeq2/GO_enrichment//" # Define the path for the directory wehre DESeq2 results will be saved

```

```{r}
# 0.4. Import the input files for the analysis

control.tpm.data <- read.csv(file=paste0(data_dir,"control_training_tpm.tsv"), sep = "\t")
View(control.tpm.data)

treatment.tpm.data <- read.csv(file=paste0(data_dir,"treatment_training_tpm.tsv"), sep = "\t")
View(treatment.tpm.data)
```


## STEP 1. FOR EACH CONDITION/TREATMENT WE FILTER THE EXPRESSED GENES BASED ON THE TPM IN ORDER TO SEE ONLY THE DIFFERENTIAL GENE EXPRESSION. EVERY CONDITION IS CONSIDERED TOGETHER

### 1.1. Create the tpm table with only gene expressed with tpm above certain value. The threshold used is tpm > 0.5

```{r}
# 1.1.1. Prepare the dataset of the first condition (here "control"). In this example we use two different time points

control.tpm.data$T3 <- (control.tpm.data[,1] + control.tpm.data[,2] + control.tpm.data[,3]) / 3 # Calculate the mean for the tpm of the replicates and generate a new column with the name after $ ("T3")
control.tpm.data$T7 <- (control.tpm.data[,4] + control.tpm.data[,5] + control.tpm.data[,6]) / 3

# Create a new table with the calculated means of tpm
colnames(control.tpm.data)[7:8] # Define the names of the columns for the new table
tpmData.average.control <- control.tpm.data[,7:8] # Create the table choosing the right columns
View(tpmData.average.control)

# In the table of means, create a new column that indicates the maximum tpm among the time points
tpmData.average.control$maxtpm <- apply(tpmData.average.control[,1:2],1,max)
View(tpmData.average.control)

# Create a new table filtering the transcripts based on the threshold maxtpm > 0.5
control_genes_0.5tpm <- tpmData.average.control[tpmData.average.control$maxtpm > 0.5,]
dim(tpmData.average.control)
dim(control_genes_0.5tpm)

View(control_genes_0.5tpm)
```

```{r}
# 1.1.2. Prepare the data set of the second condition (here "treatment"). In this example we use two different time points

treatment.tpm.data$T3 <- (treatment.tpm.data[,1] + treatment.tpm.data[,2] + treatment.tpm.data[,3]) / 3 # Calculate the mean for the tpm of the replicates and generate a new colum with the name after $ ("T3")
treatment.tpm.data$T7 <- (treatment.tpm.data[,4] + treatment.tpm.data[,5] + treatment.tpm.data[,6]) / 3

# Create a new table with the calculated means of tpm
colnames(treatment.tpm.data)[7:8] # Define the names of the columns for the new table
tpmData.average.treatment <- treatment.tpm.data[,7:8] # Create the table choosing the right columns
View(tpmData.average.treatment)

# In the table of means, create a new column that indicates the maximum tpm among the time points
tpmData.average.treatment$maxtpm <- apply(tpmData.average.treatment[,1:2],1,max)
View(tpmData.average.treatment)

# Create a new table filtering the transcripts based on the threshold maxtpm > 0.5
treatment_genes_0.5tpm <- tpmData.average.treatment[tpmData.average.treatment$maxtpm > 0.5,]
dim(tpmData.average.treatment)
dim(treatment_genes_0.5tpm)
View(treatment_genes_0.5tpm)
```

```{r}
# 1.1.3. Create a new table merging the tables for all the conditions generated above

merged_genes_0.5tpm <- merge(control_genes_0.5tpm, treatment_genes_0.5tpm, by = 0, all = T)
View(merged_genes_0.5tpm)
dim(merged_genes_0.5tpm)

colnames(merged_genes_0.5tpm)[1] <- "gene" # Give the name "gene" to the first column (colum with the gene names)
View(merged_genes_0.5tpm)

## Finally remove the rows of "LC" genes.
merged_genes_0.5tpm <- merged_genes_0.5tpm[!grepl("LC",merged_genes_0.5tpm$gene), ]
View(merged_genes_0.5tpm)
dim(merged_genes_0.5tpm)

```

### 1.2. Create the gene count table for each condition with the tpm filter

```{r}
# 1.2.1 Data set of the first condition (here "control")

control.count.data <- read.csv(file = paste0(data_dir,"control_training_counts.tsv"), sep = "\t") # Import the table of gene count from the directory of trimm results
View(control.count.data)

FLB.control.count.data <- control.count.data[,1:6]
View(FLB.control.count.data)
dim(FLB.control.count.data)

# Create the new data set applying the tpm > 0.5 table 
FLB.control.count.data_0.5tpm <- FLB.control.count.data[rownames(FLB.control.count.data) %in% merged_genes_0.5tpm$gene,]
View(FLB.control.count.data_0.5tpm)
dim(FLB.control.count.data_0.5tpm)
```

```{r}
# 1.2.1 Data set of the second condition (here "treatment")
treatment.count.data <- read.csv(file = paste0(data_dir,"treatment_training_counts.tsv"), sep = "\t") # Import the table of gene count from the directory of trimm results
View(treatment.count.data)

FLB.treatment.count.data <- treatment.count.data[,1:6]
View(FLB.treatment.count.data)
dim(FLB.treatment.count.data)

# Create the new data set applying the tpm > 0.5 table 
FLB.treatment.count.data_0.5tpm <- FLB.treatment.count.data[rownames(FLB.treatment.count.data) %in% merged_genes_0.5tpm$gene,]
View(FLB.treatment.count.data_0.5tpm)
dim(FLB.treatment.count.data_0.5tpm)
```

## STEP 2. Analyse the differential gene expression with DESeq2

### 2.1. Create the matrix for DESeq2

```{r}
# 2.1.1. Create a new data set with all the conditions with the gene counts previously filtered 

counts_for_DESeq2 <- merge(FLB.control.count.data_0.5tpm, FLB.treatment.count.data_0.5tpm, by = 0)
View(counts_for_DESeq2)
rownames(counts_for_DESeq2) <- counts_for_DESeq2[,1] # Put the names of the genes in the merged table as the rownames for the data set
View(counts_for_DESeq2)
counts_for_DESeq2 <- counts_for_DESeq2[,-1] # Remove the column with the gene names (usually the first column)
View(counts_for_DESeq2)
dim(counts_for_DESeq2)
```

```{r}
# 2.1.2. Define the groups (factors) for the analysis

colnames(counts_for_DESeq2)
timepoints <- c("FLB3_control", "FLB7_control", "FLB3_treatment", "FLB7_treatment") # Create and name the vectors needed. Name the vectors in the same order as the replicates are
CondVector <- rep(timepoints,each=3) # Indicate how many replicates are in each vector
CondVector

sampleTable <- data.frame(row.names = colnames(counts), conditions = as.factor(CondVector)) # Create the data frame for the matrix that will be used for the analysis. Define the CondVector previously created as the conditions factor for the analysis
sampleTable
View(sampleTable)

counts <- round(counts_for_DESeq2) # Create a data set with the gene counts rounded
View(counts)
```

```{r}
# 2.1.3. Create the matrix for the analysis (dds)

dds <- DESeqDataSetFromMatrix(countData = counts, colData = sampleTable, design=~conditions) 
dds

dds$conditions <- relevel(dds$conditions, "FLB3_treatment") # Define which of the conditions is the reference
dim(dds)

dds_copy <- dds # Make a copy of the matrix in case streamdown steps will go wrong
```

### 2.2. DESeq2 analysis

```{r}
# 2.2.1. Run the function

dds_analysis <- DESeq(dds)
```



```{r}
# 2.2.2. Save the results of DESeq

# Create a list with the timepoints + tissues
timepoints # Vector created before with the name of each group
time_list <- c("3", "7")

# Get data for each point control vs treatment. A loop is created based on vector "time_list"

for (i in time_list) {
  
  dds_res <- results(dds_analysis, contrast = c("conditions", paste0("FLB",i,"_control"), paste0("FLB",i,"_treatment")))
  
  # Sort results on padj
  ordered_res <- dds_res[order(dds_res$padj),] # Reorder based on the padj value
  head(ordered_res)
  tail(ordered_res)
  ordered_res_na.rm <- na.omit(ordered_res) # Remove the NA gaps
  head(ordered_res_na.rm)
  tail(ordered_res_na.rm)
  
  # Output ordered_res to csv
  
  write.csv(ordered_res_na.rm[ordered_res_na.rm$padj<0.05,],file = paste0(out_dir_DESeq2,"control_vs_treatment_results.csv")) # Save the results whose padj is lower than 0.05
  
  assign(paste("FLB",i,"control_vs_treatment.na.rm",sep=""), ordered_res_na.rm[ordered_res_na.rm$padj<0.05,]) # Important the name given here is used below
  
  print(paste("FLB",i,"control_vs_treatment.na.rm",sep=""))
  dim(assign(paste("FLB",i,"control_vs_treatment.na.rm",sep=""), ordered_res_na.rm[ordered_res_na.rm$padj<0.05,]))

}

# Create a data drame to summarize the significant differential expressed genes. Separate according to range of p-value adjusted

DE_genes <- data.frame(timepoint = numeric(),
                       DE_genes_0.05 = numeric(), upreg_0.05 = numeric(), downreg_0.05 = numeric(),
                       DE_genes_0.01 = numeric(), upreg_0.01 = numeric(), downreg_0.01 = numeric(),
                       DE_genes_0.001 = numeric(), upreg_0.001 = numeric(), downreg_0.001 = numeric())

DE_genes

for (i in time_list) { 
  
  head(get(paste("FLB",i,"control_vs_treatment.na.rm",sep="")))
  my_data <- get(paste("FLB",i,"control_vs_treatment.na.rm",sep="")) # Get the data saved above
  head(my_data)
  
  # Extract the data from the DESeq2 analysis. The range of p-value are those who where in the "DE_genes" data frame
  DE_genes_0.05 <- nrow(my_data) 
  upreg_0.05 <- nrow(my_data[my_data$log2FoldChange > 1,])
  downreg_0.05 <- nrow(my_data[my_data$log2FoldChange < -1,])
  DE_genes_0.01 <- nrow(my_data[my_data$padj < 0.01,])
  upreg_0.01 <- nrow(my_data[my_data$padj < 0.01 & my_data$log2FoldChange > 1,])
  downreg_0.01 <- nrow(my_data[my_data$padj < 0.01 & my_data$log2FoldChange < -1,])
  DE_genes_0.001 <- nrow(my_data[my_data$padj < 0.001,])
  upreg_0.001 <- nrow(my_data[my_data$padj < 0.001 & my_data$log2FoldChange > 1,])
  downreg_0.001 <- nrow(my_data[my_data$padj < 0.001 & my_data$log2FoldChange < -1,])
  
  # Fill the data frame with the data extracted
  DE_genes <- rbind(DE_genes, list(timepoint = as.numeric(i),
                                   DE_genes_0.05 = DE_genes_0.05, upreg_0.05 = upreg_0.05, downreg_0.05 = downreg_0.05,
                                   DE_genes_0.01 = DE_genes_0.01, upreg_0.01 = upreg_0.01, downreg_0.01 = downreg_0.01,
                                   DE_genes_0.001 = DE_genes_0.001, upreg_0.001 = upreg_0.001, downreg_0.001 = downreg_0.001))
}

DE_genes

colnames(DE_genes) <- c("timepoint", "DE_genes_0.05", "upreg2fold_0.05", "downreg2fold_0.05", "DE_genes_0.01", "upreg2fold_0.01", "downreg2fold_0.01", "DE_genes_0.001", "upreg2fold_0.001", "downreg2fold_0.001") # Names of the columns in the data frame. Should match the names defined above to avoid confusions

write.csv(DE_genes, file = paste0(out_dir_DESeq2,"DE_genes_results.csv")) # Save the data frame/table created with the proper name in .csv format

```

## STEP 3. GO term enrichment for each time point for the three thresholds

### 3.1. Read in information about lengths and GO terms 

```{r}
# 3.1.1. Read in GO terms

all_go <- read.csv("//jic-hpc-data/HPC-Home/rna-training/R/IWGSC_stress_GO.csv",sep = ",") # Import the GO terms file
head(all_go)
all_go <- all_go[,c(1,2)]
colnames(all_go) <- c("Gene", "GO_term") # Name properly each colum for the downstream steps
head(all_go)
dim(all_go)


```

```{r}
# 3.1.1.(Extra) For the case of "IWGSC_stress_GO.csv" file, convert the file from v1.0 to v1.1. Maybe this chunk could be unnecessary in the future.

# Note from the original script: "this is probably not the perfect thing to do so it might need to be re-done.....". In any case, this code works for the aim of the script I think

head(gsub("01G", "02G", all_go$Gene))

all_go$Gene <- (gsub("01G", "02G", all_go$Gene))
head(all_go)
dim(all_go)

all_go_HC <- all_go[!grepl("LC", all_go$Gene),] # Remove the LC genes of the GO_term v1.0
head(all_go_HC)
dim(all_go_HC)

length(unique(all_go_HC$Gene)) # Number of HC (high-confidence) genes with go terms before removing ones which don't match v1.0 to v1.1


genes_to_transfer <- read.csv(file="//jic-hpc-data/HPC-Home/rna-training/R/genes_to_transfer_qcov90_pident99_same_ID.csv") # Import the v1.1 of the GO_terms 
head(genes_to_transfer)

all_go_new <- all_go[all_go$Gene %in% genes_to_transfer$gene_v1.1,] # Only keep genes which were >99 % ID > 90% coverage from v1.0 to v1.1 
head(all_go_new)
dim(all_go_new)

length(unique(all_go_new$Gene)) # Number of genes with go terms
```
```{r}
# 3.1.2. Select those genes which were used for DESeq2 (after the filter tpm >0.5)

head(counts_for_DESeq2) 
dim(counts_for_DESeq2)

all_go_subset <- subset(all_go_new, Gene%in%rownames(counts_for_DESeq2))
View(all_go_subset)
dim(all_go_subset)

```

```{r}
# 3.1.3. Create a vector for the gene lengths. We need to get lengths of the genes not of transcripts (output from tximport script)

gene_lengths <- read.csv(file = paste0(data_dir,"control_training_gene_length.csv"), header =T) # Import the file with the tximport gene lengths
head(gene_lengths)
colnames(gene_lengths) <- c("gene", "length") # Name properly the columns of the table
head(gene_lengths)
dim(gene_lengths)

# Select those genes which were used for DESeq2, similarly to the chunk above for GO term vector
t1 <- subset(gene_lengths, gene %in% rownames(counts_for_DESeq2))
head(t1)
dim(t1)

# Create vectors that could be use with GO.Seq
gene.lens <- as.numeric(t1$length) # Vector of gene lengths for the bias.data of nullp function (below)
names(gene.lens) = t1$gene
head(gene.lens)
length(gene.lens)

assayed.genes <- as.vector(t1$gene) # Vector with (names of) the genes assayed that later on will be separate according to p-value adj from DESeq2
length(assayed.genes)

```

### 3.2. GO term enrichment. Manual: https://bioinformatics-core-shared-training.github.io/cruk-summer-school-2020/RNAseq/extended_html/06_Gene_set_testing.html

```{r}
# 3.2.1. Create the data frame for the output

GO_enriched <- data.frame(category = character(), over_represented_pvalue =  numeric(), under_represented_pvalue = numeric(), numDEInCat = numeric(), numInCat = numeric(), term = character(), ontology = character(), over_rep_padj = numeric(), timepoint_threshold = character())

head(GO_enriched)

# 3.2.2. Do the GO enrichment for each time point. The time_list object was defined in DESeq2 #2.2.2. chunk

for (i in time_list) {
  
  head(get(paste("FLB",i,"control_vs_treatment.na.rm",sep = ""))) # Define in in DESeq2 #2.2.2. chunk, for each time point in the item list
  my_data <- get(paste("FLB",i,"control_vs_treatment.na.rm",sep = "")) # Get the data from DESeq2 analysis
  head(my_data)
  
  # Create a sublist of items (j) each item is the vector with the differential expressed genes inside the range (based on the padj of DESeq2). We can choose the ranges that are more interesting
  upreg2fold_0.05 <- (my_data[my_data$padj < 0.05 & my_data$log2FoldChange > 1,])
  downreg2fold_0.05 <- (my_data[my_data$padj < 0.05 & my_data$log2FoldChange < -1,])
  upreg2fold_0.01 <- (my_data[my_data$padj < 0.01 & my_data$log2FoldChange > 1,])
  downreg2fold_0.01 <- (my_data[my_data$padj < 0.01 & my_data$log2FoldChange < -1,])
  upreg2fold_0.001 <- (my_data[my_data$padj < 0.001 & my_data$log2FoldChange > 1,])
  downreg2fold_0.001 <- (my_data[my_data$padj < 0.001 & my_data$log2FoldChange < -1,])
  
  for (j in c("upreg2fold_0.05","downreg2fold_0.05","upreg2fold_0.01","downreg2fold_0.01","upreg2fold_0.001","downreg2fold_0.001")) { # Indicate the name of the (j) items as they were defined just above
    genes_for_GO <- (get(j))
    head(genes_for_GO)
    
    # Now do GO stats analysis on the genes expressed in each pattern compared to all genes expressed
    # Create a named binary vector for genes where one means differently expressed and 0 means not differentially expressed
    de.genes <- rownames(genes_for_GO)
    gene.vector= as.integer(assayed.genes%in%de.genes)
    names(gene.vector)=assayed.genes
    head(gene.vector)
    
    # Carry out the GOseq analysis
    
    pwf = nullp(gene.vector, bias.data = gene.lens, plot.fit = TRUE)
    GO.wall = goseq(pwf,gene2cat = all_go_subset)
    # This gave table with p-values. Now correct for multiple testing using FDR
    enriched.GO = GO.wall$category[p.adjust(GO.wall$over_represented_pvalue, method = "BH")<.05]
    head(enriched.GO)
    
    # Add new column with over represented GO terms padj
    GO.wall$over_rep_padj = p.adjust(GO.wall$over_represented_pvalue, method = "BH")
    write.table(GO.wall[GO.wall$over_rep_padj<0.05,], file = paste0(out_dir_GO,i,"control_vs_treatment_results",j,"_GOseq.tsv", sep = ""), sep = "\t", quote = FALSE, col.names = TRUE, row.names = F)
    GO_enriched_timepoint <- GO.wall[GO.wall$over_rep_padj <0.05 & GO.wall$ontology == "BP",]
    head(GO_enriched_timepoint)
    
    # if no enriched GO terms don't add to dataframe
    
    if(nrow(GO_enriched_timepoint)>0) {
    
    GO_enriched_timepoint$timepoint_threshold <- paste0(i,"training_",j) # Warning: the way to call each category in this column "timepoint_threshold" will be important for one step ahead
    GO_enriched <- rbind(GO_enriched,GO_enriched_timepoint)
    
    }
}
}


```

### 3.3. Re-arrange the table to be in an easier interpretet format

```{r}
# Start from GO_enriched object

head(GO_enriched)
View(GO_enriched)

# Group based on "timepoint_threshold" categories and give ranks based on over_rep_padj result

GO_enriched_ranked <- GO_enriched %>%
  group_by(timepoint_threshold) %>%
  mutate(subrank = rank(over_rep_padj,ties.method = "first"))

# Check whether rank worked

head(data.frame(GO_enriched_ranked))
colnames(GO_enriched_ranked)
View(GO_enriched_ranked)


```

```{r}
# Now select columns to spread

GO_enriched_sel <- as.data.frame(GO_enriched_ranked[,c(6,9,10)]) # Select the columns "term", "time_threshold" and "subrank"
head(GO_enriched_sel)
View(GO_enriched_sel)

GO_data_spread <- spread(GO_enriched_sel, subrank, term) # The spread function put in the same row the terms for each timepoint_threshold
dim(GO_data_spread)
head(GO_data_spread)

unique(GO_data_spread$timepoint_threshold)
head(GO_data_spread)

```

```{r}

# Want to add number of genes in each DE group, and groups without BP GO terms enriched

DE_genes_long <- DE_genes%>% 
  pivot_longer(!timepoint, names_to = "threshold") # Create a data frame where the columns (DE_genes_0.05, upreg2fold_0.05, etc) created for DESeq2 output are now rows instead
head(DE_genes_long)
tail(DE_genes_long)

# select only DE 2 fold padj 0.05, 0.01 or 0.001 and make a new column with the "timepoint_threshold" to facilitate merging
DE_genes_long_categories <- DE_genes_long %>%
  filter(threshold == "downreg2fold_0.05" |
           threshold == "upreg2fold_0.05" |
           threshold == "downreg2fold_0.001" |
           threshold == "upreg2fold_0.001" |
           threshold == "upreg2fold_0.01" |
           threshold == "downreg2fold_0.01" ) %>%
  mutate(timepoint_threshold = paste0(timepoint,"training_",threshold)) # The word "training_" should be the same as it is indicated in the warning meassage in GOseq chunk #3.2.2.

head(DE_genes_long_categories)

```

```{r}
# Merge both tables GO_data_spread and DE_gene_long_categories

merged_GO_data_spread <- merge(DE_genes_long_categories, GO_data_spread, by = "timepoint_threshold")

View(merged_GO_data_spread)
dim(merged_GO_data_spread)

merged_GO_data_spread_by_timepoint <- merged_GO_data_spread[order(merged_GO_data_spread$timepoint),] # Short by timepoint
colnames(merged_GO_data_spread_by_timepoint)[4] <- "num_DE_genes" # Rename value column to num_DE_genes

head(merged_GO_data_spread_by_timepoint)
View(merged_GO_data_spread_by_timepoint)

```

```{r}
# Separate into 0.05, 0.01 and 0.001

merged_GO_data_spread_0.05 <- merged_GO_data_spread_by_timepoint %>%
  filter(threshold == "upreg2fold_0.05" | threshold == "downreg2fold_0.05")
head(merged_GO_data_spread_0.05)

merged_GO_data_spread_0.01 <- merged_GO_data_spread_by_timepoint %>%
  filter(threshold == "upreg2fold_0.01" | threshold == "downreg2fold_0.01")
head(merged_GO_data_spread_0.01)

merged_GO_data_spread_0.001 <- merged_GO_data_spread_by_timepoint %>%
  filter(threshold == "upreg2fold_0.001" | threshold == "downreg2fold_0.001")
head(merged_GO_data_spread_0.001)

# Save to a .csv file

write.csv(file=paste0(out_dir_GO,"GO_enrichment_padj0.05.csv"), merged_GO_data_spread_0.05, row.names = F)

write.csv(file=paste0(out_dir_GO,"GO_enrichment_padj0.01.csv"), merged_GO_data_spread_0.01, row.names = F)

write.csv(file=paste0(out_dir_GO,"GO_enrichment_padj0.001.csv"), merged_GO_data_spread_0.001, row.names = F)

```
