---
title: "GO enrichment visualization"
author: "Ismael Gutierrez-Fernandez"
date: "2025-03-27"
output: html_document
---

### The aim of this script is to plot the results of the GO term enrichement (with goseq) from the saved files. The dotplots are designed to have as much information as possible.

```{r setup, include=FALSE}
# Install packages. Only run the first time if needed.
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggpubr")
```

```{r}
# Load packages
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(dplyr)
library(forcats)
```


# Import the files from GO term enrichment at certain padj range

```{r}
# Define the path where the GO term results from goseq are saved
data_dir <- "//jic-hpc-data/Group-Scratch/Philippa-Borrill/Ismael/DESeq2/HT41x2_water_deficits/GO_enrichment/"

# Import up-regulated GO terms in LWD vs WW conditions.
GOterm.up_LWD <- read.csv(file=paste0(data_dir,"HT41_LWD_vs_WW_results_upreg2fold_0.01_GOseq.tsv"), sep = "\t")
dim(GOterm.up_LWD)

# Import down-regulated GO terms in LWD vs WW conditions.
GOterm.down_LWD <- read.csv(file=paste0(data_dir,"HT41_LWD_vs_WW_results_downreg2fold_0.01_GOseq.tsv"), sep = "\t")
dim(GOterm.down_LWD)

##################################################################################################################

# Import up-regulated GO terms in TWD vs WW conditions.
GOterm.up_TWD <- read.csv(file=paste0(data_dir,"HT41_TWD_vs_WW_results_upreg2fold_0.01_GOseq.tsv"), sep = "\t")
dim(GOterm.up_TWD)

# Import down-regulated GO terms in TWD vs WW conditions.
GOterm.down_TWD <- read.csv(file=paste0(data_dir,"HT41_TWD_vs_WW_results_downreg2fold_0.01_GOseq.tsv"), sep = "\t")
dim(GOterm.down_TWD)

```

## PLOTS OF LONG WATER DEFICIT (LWD) VS WW CONDITIONS.

### UP-REGULATED

```{r}
# Select the GO terms based on their ontology: biological process (BP), molecular function (MF), and cellular components (CC). For the aim of the study, I focus on BP and MF.

GOterm.up_LWD_BP <-  filter(GOterm.up_LWD, GOterm.up_LWD$ontology %in% c("BP"))
dim(GOterm.up_LWD_BP)

GOterm.up_LWD_MF <-  filter(GOterm.up_LWD, GOterm.up_LWD$ontology %in% c("MF"))
dim(GOterm.up_LWD_MF)

# Plot the dot plots with each subset
# 1. Biological processes

GOterm.up_LWD_BP %>%
  top_n(20, wt = -over_represented_pvalue) %>%  # Select the top20 GO term with highest significance (lowest over_represented_pvalue)
  mutate(hitsPerc = numDEInCat*100/numInCat) %>% # Calculate the ratio of counted DEGs for that GO term and all genes associated to that GO term.
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>% # Reorder the GO terms in order to appear in descendant order based on lowaest p-value
  ggplot(aes(x = hitsPerc, # Variable in X axis
             y = term, # Variable in Y axis
             colour = -over_represented_pvalue, # Variable that define color scale
             size = numDEInCat)) + # Variable that define dots size
  geom_point(stroke = 2) + # Dotplot
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) + # To add the segment lined the dot with each GO term
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") + # Define the name of the axis and the colour and size legends
  ggtitle("Biological process, up-regulated in HT41 under LWD") + # Title of the plot
  scale_color_gradient(low = "#e69394", high = "#780002", space = "Lab") + # Define the colour scale
  theme_bw() + # White background and more aesthetic
  theme(axis.text.x = element_text(size=rel(1.15)), 
        axis.title = element_text(size=rel(1.15))) + # Define the size of the axis title
  scale_x_continuous(limits = c(0, 100)) # Modify the values showed in the axis X that could be adjusted for each GO term plot


# 2. Molecular function
GOterm.up_LWD_MF %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Molecular function, up-regulated in HT41 under LWD") +
  scale_color_gradient(low = "#e69394", high = "#780002", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))

```

### DOWN-REGULATED

```{r}
# Select the GO terms based on their ontology: biological process (BP), molecular function (MF), and cellular components (CC). For the aim of the study, I focus on BP and MF.
GOterm.down_LWD_BP <-  filter(GOterm.down_LWD, GOterm.down_LWD$ontology %in% c("BP"))
dim(GOterm.down_LWD_BP)

GOterm.down_LWD_MF <-  filter(GOterm.down_LWD, GOterm.down_LWD$ontology %in% c("MF"))
dim(GOterm.down_LWD_MF)

# Plot the dot plots with each subset
# 1. Biological processes
GOterm.down_LWD_BP %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Biological proccess, down-regulated in HT41 under LWD") +
  scale_color_gradient(low = "#add2ff", high = "#0a63cf", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))

# 2. Molecular function
GOterm.down_LWD_MF %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Molecular function, down-regulated in HT41 under LWD") +
  scale_color_gradient(low = "#add2ff", high = "#0a63cf", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))

```


## PLOTS OF TERMINAL WATER DEFICIT (TWD) VS WW CONDITIONS.

### UP-REGULATED

```{r}
# Select the GO terms based on their ontology: biological process (BP), molecular function (MF), and cellular components (CC). For the aim of the study, I focus on BP and MF.

GOterm.up_TWD_BP <-  filter(GOterm.up_TWD, GOterm.up_TWD$ontology %in% c("BP"))
dim(GOterm.up_TWD_BP)

GOterm.up_TWD_MF <-  filter(GOterm.up_TWD, GOterm.up_TWD$ontology %in% c("MF"))
dim(GOterm.up_TWD_MF)

# Plot the dot plots with each subset
# 1. Biological processes

GOterm.up_TWD_BP %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Biological proccess, up-regulated in HT41 under TWD") +
  scale_color_gradient(low = "#e69394", high = "#780002", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))


# 2. Molecular function
GOterm.up_TWD_MF %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Molecular function, up-regulated in HT41 under TWD") +
  scale_color_gradient(low = "#e69394", high = "#780002", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))
```

### DOWN-REGULATED

```{r}
# Select the GO terms based on their ontology: biological process (BP), molecular function (MF), and cellular components (CC). For the aim of the study, I focus on BP and MF.
GOterm.down_TWD_BP <-  filter(GOterm.down_TWD, GOterm.down_TWD$ontology %in% c("BP"))
dim(GOterm.down_TWD_BP)

GOterm.down_TWD_MF <-  filter(GOterm.down_TWD, GOterm.down_TWD$ontology %in% c("MF"))
dim(GOterm.down_TWD_MF)

# Plot the dot plots with each subset
# 1. Biological processes
GOterm.down_TWD_BP %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Biological proccess, down-regulated in HT41 under TWD") +
  scale_color_gradient(low = "#add2ff", high = "#0a63cf", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))

# 2. Molecular function
GOterm.down_TWD_MF %>%
  top_n(20, wt = -over_represented_pvalue) %>% 
  mutate(hitsPerc = numDEInCat*100/numInCat) %>%
  mutate(term = fct_reorder(term, desc(over_represented_pvalue))) %>%
  ggplot(aes(x = hitsPerc, 
             y = term, 
             colour = -over_represented_pvalue, 
             size = numDEInCat)) +
  geom_point(stroke = 2) +
  geom_segment(aes(xend = 0, yend = term), linewidth = 0.5, linetype = 2) +
  expand_limits(x=0) +
  labs(x = "Gene ratio (%)", y = "GO term", colour = "p-value", size = "Count") +
  ggtitle("Molecular function, down-regulated in HT41 under TWD") +
  scale_color_gradient(low = "#add2ff", high = "#0a63cf", space = "Lab") +
  theme_bw() +
  theme(axis.text.x = element_text(size=rel(1.15)),
        axis.title = element_text(size=rel(1.15))) +
  scale_x_continuous(limits = c(0, 100))
```
